---
title: "A Look At Daily Returns of the S&P 500"
author: "Rex Macey"
date: "December 6, 2015"
output: html_document
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
setwd("~/Quant Trading/sp_daily")
library(quantmod)
library(TTR)
library(xts)
library(timeDate)
library(FinCal)
library(RQuantLib)
library(lubridate)


tradingDayOfMonth <- function(myDate, calendar = "UnitedStates/NYSE") {
    FirstOfMonth <- as.Date(paste(year(myDate), month(myDate), "01", sep="/")) 
    businessDaysBetween(calendar, from = FirstOfMonth, to = myDate,
                        includeFirst = 1, includeLast = 1)+1
}
tradingDaysLeftInMonth <- function(myDate, calendar = "UnitedStates/NYSE") {
    EOM <- getEndOfMonth(calendar = "UnitedStates/NYSE",myDate)
    businessDaysBetween(calendar, from = myDate, to = EOM,
                        includeFirst = 1, includeLast = 1)+1
}
DaysFromLastHoliday <- function(myDate, hdays,calendar = "UnitedStates/NYSE"){
    d <- hdays[hdays<myDate]
    businessDaysBetween(calendar, from = d[which.min(abs(myDate-d))], to = myDate,
                        includeFirst = 1, includeLast = F)
}
DaysTilNextHoliday <- function(myDate, hdays,calendar = "UnitedStates/NYSE"){
    d <- hdays[hdays>myDate]
    businessDaysBetween(calendar, from = myDate, to = d[which.min(abs(myDate-d))],
                        includeFirst = 1, includeLast = F)
}
ReplwNA <- function(x, threshold){ # replace a vector with NA if value exceeds threshold
    x[x>threshold]<-NA
    return(x)
}

spx<-readRDS("spx.rds")
rownames(spx)<-spx$Date
spx$Date=NULL
spx<-as.quantmod.OHLC(spx,c("Open","High","Low","Close","Volume","Adjusted"),name="spx")
colnames(spx)<-c("Open","High","Low","Close","Volume","Adjusted")
spx.ret<-xts(spx, order.by = as.Date(index(spx)))
spx.ret<-dailyReturn(spx.ret,leading=FALSE)
if (TRUE){
    data<-spx[,"Close",drop=FALSE]
    colnames(data)<-"Close"
    data$SMA2<-data$Close/SMA(data$Close,2) # Close relative to simple moving average
    data$SMA5<-data$Close/SMA(data$Close,5)
    data$SMA10<-data$Close/SMA(data$Close,10)
    data$SMA20<-data$Close/SMA(data$Close,20)
    data$SMA50<-data$Close/SMA(data$Close,50)
    data$SMA200<-data$Close/SMA(data$Close,200)
    data$SMA50_200<-SMA(data$Close,50) - SMA(data$Close,200) # 50d SMA -  200d SMA
    data$BB<-BBands(data$Close)[,4] # %B of Bollinger Band
    data$wday<-factor(weekdays(index(data),TRUE)) # day of week
    data$tdayofmonth<-sapply(index(data),tradingDayOfMonth) # trading day of month
    data$tdaysleft<-sapply(index(data),tradingDaysLeftInMonth) # trading day remaining in month
    data$month<-factor(month(index(data))) # month
    data$vol5<-rollapply(data.frame(spx.ret),5,sd,fill=NA) # 5 day vol
    data$vol10<-rollapply(data.frame(spx.ret),10,sd,fill=NA) #10 day vol
    data$volchg=data$vol5-data$vol10 # 5d vol - 10d vol (change in vol)
    NYSEhdays<-as.Date(holidayNYSE(year(index(data)[1]):(year(index(data)[nrow(data)])+1)))
    data$daysfromhday <- sapply(index(data),DaysFromLastHoliday,hdays=NYSEhdays)
    data$daystilhday <- sapply(index(data),DaysTilNextHoliday,hdays=NYSEhdays)
    data$SMA50_200Factor<-data[,"SMA50_200"]>=0
    data$ret <- spx.ret
    saveRDS(data,"data.rds")
} else {
    data<-readRDS("data.rds")
}
data[,"tdayofmonth"]<-ReplwNA(data[,"tdayofmonth"],10)
data[,"tdaysleft"]<-ReplwNA(data[,"tdaysleft"],10)
data[,"daysfromhday"]<-ReplwNA(data[,"daysfromhday"],10)
data[,"daystilhday"]<-ReplwNA(data[,"daystilhday"],10)
```
Let's look at daily returns of the S&P 500. Here we have daily returns on the S&P 500 from `r format(index(spx[1]),"%B %d, %Y")` through `r format(index(spx[nrow(spx)]),"%B %d, %Y")`.  Note that the early data effectively only has closing prices. More recent data has open/high/low and close.  There are `r format(nrow(spx),nsmall=0,width=5)` prices (and `r format(nrow(spx)-1,nsmall=0,width=5)` returns)
```{r}
head(spx)
tail(spx)
```
```{r echo=FALSE}
plot(spx$Close,log="y",main="Daily Value of S&P 500 Index", ylab="Close",xlab="",col="blue")
```
Let's look at the distribution of daily returns.  We see that the worst daily return was `r paste0(round(100*min(spx.ret,na.rm=T),2),"%")`. This occurred on `r format(index(spx.ret)[which.min(spx.ret)],"%B %d, %Y")`.  The best return was `r paste0(round(100*max(spx.ret,na.rm=T),2),"%")` which occurred on `r format(index(spx.ret)[which.max(spx.ret)],"%B %d, %Y")`.
```{r echo=FALSE}
summary(spx.ret)
```
A histogram shows the vast majority are clustered near 0%, but the scale indicates there are rare, large outliers.  Recall there are `r format(nrow(spx)-1,nsmall=0,width=5)` daily returns in our dataset.

```{r echo=FALSE}
hist(spx.ret, main="Histogram of Daily Returns",xlab="",ylab="Frequency")
```

```{r echo=FALSE}
par(mfrow=c(1,2))
b1<-boxplot(coredata(spx.ret), main="Boxplot of Daily Returns w Outliers",xlab="",ylab="Return",col="blue",notch=F, outline=T,
        pars=list(boxwex=1.5))
boxplot(coredata(spx.ret), main="Boxplot of Daily Returns w/o Outliers",xlab="",ylab="Return",col="blue",notch=F, outline=F,
        pars=list(boxwex=1.5))
par(mfrow=c(1,1))
```

The boxplots show the dispersion as well.  The blue boxes contain one-half of the observations.  The whiskers (horizontal lines outside the blue) are 1.5x the width of the boxes.  Points beyond the whiskers are outliers. About `r paste0(round(100*length(b1$out)/b1$n,1),"%")` of the data are outliers using this definition.

# Can we do better than average?

Again the average and median daily returns are `r paste0(round(100*mean(spx.ret,na.rm=T),4),"%")` and `r paste0(round(100*median(spx.ret,na.rm=T),4),"%")` respectively. An investment of $1 would have grown to `r paste0("$",round(prod(1+spx.ret,na.rm=TRUE),2))` ignoring taxes and dividends.   The question is can we identify type of days which might be better than others?

```{r echo=FALSE}

```
Let's start with an example called the turn of the month anomoly.  With this we will only invest for the 3 trading days before and after the turn of the month.  

```{r echo=FALSE}
g<-data[,"tdayofmonth"]<=3 | data[,"tdaysleft"]<=3

par(mfrow=c(1,2))
b1<-boxplot(coredata(spx.ret), main="All Daily Returns",xlab="",ylab="Return",col="blue",notch=F, outline=F,
        pars=list(boxwex=1.5), ylim=c(b1$stats[1],b1$stats[5]))
lab.y<-c(b1$stats[4]+.75*(b1$stats[4]-b1$stats[2]),
         b1$stats[4]+.50*(b1$stats[4]-b1$stats[2]),
         b1$stats[4]+.25*(b1$stats[4]-b1$stats[2]))
lab.text<-c(paste0("Mean=",round(100*mean(spx.ret,na.rm=T),4),"%"),
            paste0("Median=",round(100*median(spx.ret,na.rm=T),4),"%"),
            paste0("#obs=",sum(!is.na(spx.ret))))
text(.75,lab.y,labels=lab.text)
b2<-boxplot(coredata(spx.ret[g]), main="'Turn of the Month' Returns",xlab="",ylab="Return",col="green",notch=F, outline=F,
        pars=list(boxwex=1.5), ylim=c(b1$stats[1],b1$stats[5]))
lab.text<-c(paste0("Mean=",round(100*mean(spx.ret[g],na.rm=T),4),"%"),
            paste0("Median=",round(100*median(spx.ret[g],na.rm=T),4),"%"),
            paste0("#obs=",sum(!is.na(spx.ret[g]))))
text(.75,lab.y,labels=lab.text)
z<-(mean(spx.ret[g],na.rm=T)-mean(spx.ret,na.rm=T))/(sd(spx.ret[g],na.rm=T)/sqrt(sum(!is.na(spx.ret[g]))))
p<-pnorm(z)
lab.text<-c(paste0("     z=",format(round(z,3),nsmall=3,width=6,digits=5)),
            paste0("pvalue=",format(round(1-p,4),nsmall=4,width=6,digits=5,scientific = F)),"")
text(1.25,lab.y,labels=lab.text)
abline(h=b1$stats[3],col="blue")
abline(h=b1$stats[2],col="blue", lty=2)
abline(h=b1$stats[4],col="blue", lty=2)
par(mfrow=c(1,1))
```

The average return around the turn of the month is significantly different from the average daily return.  In fact it is about `r paste0(round(mean(spx.ret[g],na.rm=T)/mean(spx.ret,na.rm=T),1),"x")` higher.  That said, these only account for `r paste0(round(100*sum(g,na.rm=T)/sum(!is.na(spx.ret)),1),"%")` of the observations.  If one only invested on these days, $1 would have grown into `r paste0("$",round(prod(1+spx.ret[g],na.rm=T),2))`.   

```{r echo=FALSE}
sumnotisna<-function(x){
    sum(!is.na(x))
}
fld.stats<-function(stat.params,data){
    fld<-stat.params$field
    desc<-stat.params$desc
    df<-data.frame(data[,c(fld,"ret")])
    out<-list()
    out$popmean<-mean(df$ret,na.rm=T)
    out$popsd<-sd(df$ret,na.rm=T)
    agg.ret<-aggregate(df,by=list(df[,fld]),FUN=mean,na.rm=T)
    agg.n<-aggregate(df,by=list(df[,fld]),FUN=sumnotisna)
    agg.sd<-aggregate(df,by=list(df[,fld]),FUN=sd,na.rm=T)
    agg.z<-(agg.ret$ret-out$popmean)/(agg.sd$ret/sqrt(agg.n$ret))
    agg.p<-1-pnorm(agg.ret$ret,mean=out$popmean,sd=agg.sd$ret/sqrt(agg.n$ret))
    agg<-data.frame(ret=agg.ret$ret)
    agg$n<-agg.n$ret
    agg$sd<-agg.sd$ret
    agg$sd.up<-agg$sd*sqrt((agg$n-1)/qchisq(.025,agg$n-1))
    agg$sd.dn<-agg$sd*sqrt((agg$n-1)/qchisq(1-.025,agg$n-1))
    agg$z<-agg.z
    agg$p<-agg.p
    out$agg<-agg
    barplot(out$agg$ret,names.arg=row.names(out$agg),
                  main=paste("Average returns by",desc),
                  col="blue",ylab="Mean Return",xlab=desc)
    abline(h=out$popmean,col="red")
    out$box1<-boxplot(as.formula(paste0("ret~",fld)),data=df,
                      main=paste("Returns by",desc),
                      col="blue",ylab="Return",xlab=desc,
                      names.arg=row.names(out$agg))
    out$box2<-boxplot(as.formula(paste0("ret~",fld)),data=df,outline=F,
                      main=paste("Returns by",desc),
                      col="blue",ylab="Return",xlab=desc,
                      names.arg=row.names(out$agg))
    abline(h=out$popmean,col="red")
    out$bar2<-barplot(out$agg$z,col="blue",names.arg=row.names(out$agg),
                      main=paste("z-values by",desc),ylab="z value",xlab=desc,
                      sub="z value is (avgret - popmean)/(sd/sqrt(n))",ylim=c(-3,3))
    abline(h=qnorm(0.95),col="gray",lty=2)
    abline(h=qnorm(0.99),col="gray",lty=3)
    abline(h=qnorm(0.05),col="gray",lty=2)
    abline(h=qnorm(0.01),col="gray",lty=3)
    text(x=nrow(out$agg)/2,y=c(qnorm(0.01),qnorm(0.05),qnorm(0.95),qnorm(0.99)),
         labels=c("1% Probability","5% Probability","95% Probability","99% Probability"))
    
    out$sdplot<-barplot(out$agg$sd,col="blue",
                        main=paste("Standard Deviations by ",desc),
                        ylab="Standard Deviation",
                        xlab=desc,names.arg=row.names(out$agg))
    segments(out$sdplot,out$agg$sd.dn,out$sdplot,out$agg$sd.up,col="green",lwd=6)
    epsilon=0.25
    segments(out$sdplot-epsilon,out$agg$sd.dn,
             out$sdplot+epsilon,out$agg$sd.dn,col="green",lwd=3)
    segments(out$sdplot-epsilon,out$agg$sd.up,
             out$sdplot+epsilon,out$agg$sd.up,col="green",lwd=3)
    abline(h=out$popsd,col="red")
}
```



# Calendar Month
```{r echo=FALSE}
stat.params<-list(field="month",desc="Calendar Month")
print(fld.stats(stat.params,data))
```

# Trading Day of Month
```{r echo=FALSE}
stat.params<-list(field="tdayofmonth",desc="Trading Day of Month")
print(fld.stats(stat.params,data))
```

# Trading Days Left in Month
```{r echo=FALSE}
stat.params<-list(field="tdaysleft",desc="Trading Days Left in Month")
print(fld.stats(stat.params,data))
```

# Day of Week
```{r echo=FALSE}
stat.params<-list(field="wday",desc="Weekday")
print(fld.stats(stat.params,data))
```

# Days from Holiday
```{r echo=FALSE}
stat.params<-list(field="daysfromhday",desc="Days from Holiday")
print(fld.stats(stat.params,data))
```

# Days until Holiday
```{r echo=FALSE}
stat.params<-list(field="daystilhday",desc="Days til Holiday")
print(fld.stats(stat.params,data))
```

# SMA2 - Price / 2 day simple moving average
```{r echo=FALSE}
sma2_qtile<-cut(data[,"SMA2"],quantile(data[,"SMA2"],c(0,.2,.4,.6,.8,1),na.rm=TRUE))
data<-cbind(data,sma2_qtile)
stat.params<-list(field="sma2_qtile",desc="Quintile of Price/SMA2")
print(fld.stats(stat.params,data))
```

# SMA5 - Price / 5 day simple moving average
```{r echo=FALSE}
sma5_qtile<-cut(data[,"SMA5"],quantile(data[,"SMA5"],c(0,.2,.4,.6,.8,1),na.rm=TRUE))
data<-cbind(data,sma5_qtile)
stat.params<-list(field="sma5_qtile",desc="Quintile of Price/SMA5")
print(fld.stats(stat.params,data))
```

# SMA10 - Price / 10 day simple moving average
```{r echo=FALSE}
sma10_qtile<-cut(data[,"SMA10"],quantile(data[,"SMA10"],c(0,.2,.4,.6,.8,1),na.rm=TRUE))
data<-cbind(data,sma10_qtile)
stat.params<-list(field="sma10_qtile",desc="Quintile of Price/SMA10")
print(fld.stats(stat.params,data))
```

# SMA20 - 20 day simple moving average
```{r echo=FALSE}
sma20_qtile<-cut(data[,"SMA20"],quantile(data[,"SMA20"],c(0,.2,.4,.6,.8,1),na.rm=TRUE))
data<-cbind(data,sma20_qtile)
stat.params<-list(field="sma20_qtile",desc="Quintile of Price/SMA20")
print(fld.stats(stat.params,data))
```

# SMA50 - 50 day simple moving average
```{r echo=FALSE}
sma50_qtile<-cut(data[,"SMA50"],quantile(data[,"SMA50"],c(0,.2,.4,.6,.8,1),na.rm=TRUE))
data<-cbind(data,sma50_qtile)
stat.params<-list(field="sma50_qtile",desc="Quintile of Price/SMA50")
print(fld.stats(stat.params,data))
```

# SMA200 - 200 day simple moving average
```{r echo=FALSE}
sma200_qtile<-cut(data[,"SMA200"],quantile(data[,"SMA200"],c(0,.2,.4,.6,.8,1),na.rm=TRUE))
data<-cbind(data,sma200_qtile)
stat.params<-list(field="sma200_qtile",desc="Quintile of Price/SMA200")
print(fld.stats(stat.params,data))
```

# SMA50_200 - 50d SMA - 200d SMA
```{r echo=FALSE}
sma50_200_qtile<-cut(data[,"SMA50_200"],quantile(data[,"SMA50_200"],c(0,.2,.4,.6,.8,1),na.rm=TRUE))
data<-cbind(data,sma50_200_qtile)
stat.params<-list(field="sma50_200_qtile",desc="Quintile of SMA50 - SMA200")
print(fld.stats(stat.params,data))
```

# BB  - Bollinger Band 
```{r echo=FALSE}
bb_qtile<-cut(data[,"BB"],quantile(data[,"BB"],c(0,.2,.4,.6,.8,1),na.rm=TRUE))
data<-cbind(data,bb_qtile)
stat.params<-list(field="bb_qtile",desc="Quintile of Bollinger Band")
print(fld.stats(stat.params,data))
```

# vol5 - 5 day Volatility
```{r echo=FALSE}
vol5_qtile<-cut(data[,"vol5"],quantile(data[,"vol5"],c(0,.2,.4,.6,.8,1),na.rm=TRUE))
data<-cbind(data,vol5_qtile)
stat.params<-list(field="vol5_qtile",desc="Quintile of 5d Volatility")
print(fld.stats(stat.params,data))
```
# vol10 - 10 day Volatility
```{r echo=FALSE}
vol10_qtile<-cut(data[,"vol10"],quantile(data[,"vol10"],c(0,.2,.4,.6,.8,1),na.rm=TRUE))
data<-cbind(data,vol10_qtile)
stat.params<-list(field="vol10_qtile",desc="Quintile of 10d Volatility")
print(fld.stats(stat.params,data))
```
# volchg - 5 day - 10 day volatility 
```{r echo=FALSE}
volchg_qtile<-cut(data[,"volchg"],quantile(data[,"volchg"],c(0,.2,.4,.6,.8,1),na.rm=TRUE))
data<-cbind(data,volchg_qtile)
stat.params<-list(field="volchg_qtile",desc="Quintile of 5d-10d Volatility Change")
print(fld.stats(stat.params,data))
```